### 2016-06-29 responseToReviewers.Rmd

> _\textcolor{blue}{Although the full phrase corresponding to the acronym
in question is provided prior to the acronym in the text (along with an
explanatory reference), we have added the following clarification:}_

> \textcolor{blue}{Further, volume of FLAIR lesions within the corpus
callosum, brainstem, and thalamus in patients with severe TBI correlates
with Glasgow Outcome-Extended (GOS-E) scores} [@Moen:2014aa]
> \textcolor{blue}{---a numeric groupwise assessment used to
classify ``outcome'' in TBI patients where ``outcome'' refers to the
spectrum of possible prognoses from death to disability to recovery.}

9. In pg. 3 the authors propose that automated methods ``may allow for
identification of correlative patterns between WMH number, volume,
distribution, and disease state''. Some of these correlations have been
already identified in the literature, as the authors mention few lines
before. Moreover, the authors do not report number, volume, or
distribution of the lesions, to back up the utility of their method in
this regard.

> _\textcolor{blue}{We removed the sentence in question.  Also, as mentioned in
response to item 3, we added Table 1.}_

@@

14. The discussion starts with an ``intent to further refine the proposed
paradigm,'' and proceeds by mentioning other work currently being
performed which looks promising. The opening of the discussion with such
statements truly make the manuscript sound like an incomplete effort that
is not ready for publication.

> _\textcolor{blue}{We appreciate the positive view of our current work.
The full sentence in question is ``Although only a
single expert was used to produce the manual labelings, our intent is to
further refine the proposed paradigm by crowdsourcing with feedback from
other experts who interact with both the data and methodology.''
In our opinion, simply mentioning specific details regarding future work
does not mean that the manuscript ``sounds like an incomplete effort.''
Rather, it is simply an admission, similar to much of what is reported
in the literature, that our research program is ongoing.}_

15. The discussion also spends several paragraphs on defining literature that
shows how WMHs are related to cognitive performance, or how WMHs can be
applied in diagnosing TBI. Yet, this study makes no effort to test these
efforts in these 24 patients. Consequently, those paragraphs sound
wishful and speculative. Given the methodological nature of this paper, a
more appropriate discussion would perhaps focus on the comparison with
other methods in terms of accuracy, similarities, differences, etc.

> _\textcolor{red}{We removed the paragraphs in question and added the following
comparison discussion:}_

### 2016-06-29 responseToReviewers.tex

\begin{quote}
\emph{\textcolor{blue}{Although the full phrase corresponding to the acronym
in question is provided prior to the acronym in the text (along with an
explanatory reference), we have added the following clarification:}}
\end{quote}

\begin{quote}
\textcolor{blue}{Further, volume of FLAIR lesions within the corpus
callosum, brainstem, and thalamus in patients with severe TBI correlates
with Glasgow Outcome-Extended (GOS-E) scores}\textsuperscript{14}
\textcolor{blue}{---a numeric groupwise assessment used to
classify ``outcome'' in TBI patients where ``outcome'' refers to the
spectrum of possible prognoses from death to disability to recovery.}
\end{quote}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{8}
\tightlist
\item
  In pg. 3 the authors propose that automated methods ``may allow for
  identification of correlative patterns between WMH number, volume,
  distribution, and disease state''. Some of these correlations have
  been already identified in the literature, as the authors mention few
  lines before. Moreover, the authors do not report number, volume, or
  distribution of the lesions, to back up the utility of their method in
  this regard.
\end{enumerate}

\begin{quote}
\emph{\textcolor{blue}{We removed the sentence in question.  Also, as mentioned in
response to item 3, we added Table 1.}}
\end{quote}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{13}
\tightlist
\item
  The discussion starts with an ``intent to further refine the proposed
  paradigm,'' and proceeds by mentioning other work currently being
  performed which looks promising. The opening of the discussion with
  such statements truly make the manuscript sound like an incomplete
  effort that is not ready for publication.
\end{enumerate}

\begin{quote}
\emph{\textcolor{blue}{We appreciate the positive view of our current work.
The full sentence in question is ``Although only a
single expert was used to produce the manual labelings, our intent is to
further refine the proposed paradigm by crowdsourcing with feedback from
other experts who interact with both the data and methodology.''
In our opinion, simply mentioning specific details regarding future work
does not mean that the manuscript ``sounds like an incomplete effort.''
Rather, it is simply an admission, similar to much of what is reported
in the literature, that our research program is ongoing.}}
\end{quote}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{14}
\tightlist
\item
  The discussion also spends several paragraphs on defining literature
  that shows how WMHs are related to cognitive performance, or how WMHs
  can be applied in diagnosing TBI. Yet, this study makes no effort to
  test these efforts in these 24 patients. Consequently, those
  paragraphs sound wishful and speculative. Given the methodological
  nature of this paper, a more appropriate discussion would perhaps
  focus on the comparison with other methods in terms of accuracy,
  similarities, differences, etc.
\end{enumerate}

\begin{quote}
\emph{\textcolor{red}{We removed the paragraphs in question and added the following
comparison discussion:}}
\end{quote}

### 2016-06-29 discussion.Rmd
ENH:  Evaluation results.

Two major meta-analyses of WMHs have been published covering the periods
prior to 2010 [@Debette:2010aa] and after [@Kloppenborg:2014aa]. The earlier meta-analysis
covered 53 longitudinal studies that included samples of high-risk
populations, i.e., patients selected for a specific disease or condition
such as hypertension, whereas other studies recruited samples of the
general population. Longitudinal studies of samples representative of
the general population are more relevant to the focus of the present
paper. Debette & Markus [@Debette:2010aa] found that the presence of
WMHs was related to subsequent cognitive decline, a higher risk of
developing dementia, stroke, and of mortality. Lesion volume at baseline
was also predictive of cognitive decline. Limitations of this
meta-analysis include heterogeneity in the method of measuring WMHs;
some studies used automated volumetric measurement, whereas others used
a visual rating scale. The studies analyzed by Debette & Markus were
limited to the occurrence of one of the aforementioned conditions which
they analyzed by hazard ratios.

The more recent meta-analysis by Kloppenborg et al.
[@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and
concurrent neuropsychological results in patients with heterogeneous
diagnoses but without previously diagnosed cognitive impairment, found
that WMHs were associated with cognitive deficit (effect size of -0.10,
95% CI: -0.13 to -0.08) after controlling for age. These studies also
differed in the metric used to measure the WMHs, including volume, % of
total intracranial volume, and a visual rating score. The effect size
for the association with cognitive deficit in these cross-sectional
studies did not differ significantly across various cognitive domains or
the method of measuring lesion volume.  Among eight longitudinal studies
analyzed by Kloppenborg et al that included a follow-up MRI and also
controlled for age, the effect size for the association of progression
in WMHs and cognitive impairment was -0.16 (95% CI:-0.27 to -0.09). This
association was stronger for attention and executive function than for
memory and processing speed. Although baseline WMHs were predictive of
cognitive deficit at follow-up in the seven studies which did not repeat
MRI, the effect size was smaller [-0.10 (95% CI: 0.13 to -0.05) than in
the longitudinal studies that calculated progression in WMHs. In
summary, progression of WMHs seen on repeat MRI has a stronger relation
to cognitive deficit than concurrent imaging findings. These
meta-analyses support the rationale for repeating an MRI in patients
younger than 50 years whose initial scan shows WMHs.

Despite the
above-described associations between WMHs, cognitive decline, increased
risk of developing dementia, and mortality, these lesions receive little
attention in current clinical workflows. When reported in a standard
neuroradiologist interpretation, they are typically handled as
incidental findings and are assigned little clinical significance. This
likely reflects the impracticality of performing a detailed assessment
of number, volume, and distribution within a qualitative
neuroradiologist interpretation as well as the lack of correlative
information on how the presence and distribution of these lesions may
inform a diagnosis and prognosis in the appropriate clinical setting. To
date, automated or semi-automated tools for the detection of WMHs have
lacked the specificity and efficiency for the mining of large-scale
datasets to generate highly granular data on whether these lesions
possess any true diagnostic or prognostic value in the setting of a
specific disease process. The present communication describes a
supervised statistical learning tool that is appropriate for the
application to such large-scale datasets.

>>>>

Two major meta-analyses of WMHs have been published covering the periods
prior to [@Debette:2010aa] and after 2010 [@Kloppenborg:2014aa].
Debette & Markus [@Debette:2010aa] found that the presence of
WMHs was related to subsequent cognitive decline, a higher risk of
developing dementia, stroke, and of mortality. Lesion volume at baseline
was also predictive of cognitive decline. Kloppenborg et al.
[@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and
concurrent neuropsychological results in patients with heterogeneous
diagnoses but without previously diagnosed cognitive impairment, found
that WMHs were associated with cognitive deficit (effect size of -0.10,
95% CI: -0.13 to -0.08) after controlling for age.

Despite the potential clinical significance of WMHs
these lesions receive little
attention in current clinical workflows. When reported in a standard
neuroradiologist interpretation, they are typically handled as
incidental findings and are assigned little clinical significance. This
likely reflects the impracticality of performing a detailed assessment
of number, volume, and distribution within a qualitative
neuroradiologist interpretation as well as the lack of correlative
information on how the presence and distribution of these lesions may
inform a diagnosis and prognosis in the appropriate clinical setting. To
date, automated or semi-automated tools for the detection of WMHs have
lacked the specificity and efficiency for the mining of large-scale
datasets to generate highly granular data on whether these lesions
possess any true diagnostic or prognostic value in the setting of a
specific disease process. The present communication describes a
supervised statistical learning tool that is appropriate for the
application to such large-scale datasets.

### 2016-06-29 format.Rmd
ENH:  Evaluation results.

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

>>>>

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
  html_document:
    toc: false

header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

### 2016-06-29 intro.Rmd
ENH:  Evaluation results.

\textcolor{blue}{Despite the above findings which demonstrate that WMHs have potential prognostic value, they are not routinely employed as a diagnostic measure in clinical practice.  Performing a comprehensive manual counting of number and distribution of lesions in the clinical setting is simply not practical. As such, the development of automated methods for the rapid identification and quantification of WMHs within individual patients may allow for identification of correlative patterns between WMH number, volume, distribution, and disease state. Further, the development of such lesion quantification approaches may allow for the practical inclusion of this type of information within routine radiological practice.  In this work, we present an automated framework for quantification of WMHs in
multi-modal MRI using the random forest machine learning technique.}

>>>>

\textcolor{blue}{Despite the above findings which demonstrate that WMHs have potential prognostic value, they are not routinely employed as a diagnostic measure in clinical practice.  Performing a comprehensive manual counting of number and distribution of lesions in the clinical setting is simply not practical. The development of such lesion quantification approaches may allow for the practical inclusion of this type of information within routine radiological practice.  In this work, we present an automated framework for quantification of WMHs in
multi-modal MRI using the random forest machine learning technique.}

### 2016-06-29 methods.Rmd
ENH:  Evaluation results.

>>>>

\textcolor{blue}{Several measures have been employed in the literature for evaluating
automated white matter lesion segmentation involving such quantities as lesion load,
voxel-based overlap measures (such as the Dice similarity coefficient), and lesion-based measures.}[@Garcia-Lorenzo:2013aa]\textcolor{blue}{.  For this work, due to the relatively small-size distribution}[^4] \textcolor{blue}{of the lesion load in our data cohort (see Table 1), we used
four lesion-based measures:  sensitivity, positive predictive value (i.e., precision), $F_1$ score, and relative volume difference.  The first three quantities are based on the number of false positives ($TN$, false negatives ($FN$), and true positives ($TP$) in terms of identified lesions.  It should be noted that the number of true negatives is not readily
incorporated into measures of accuracy as the quantity of ``true negatives'' (i.e., correctly identified normal brain tissue) would severely skew accuracy assessments. The $F_1$ score is an assessment of accuracy which takes into account both the sensitivity,}

$$ sensitivity =  \frac{ TP }{TP + FN}, $$

\textcolor{blue}{and the positive predictive value (PPV),}

$$ PPV =  \frac{ TP }{TP + FP}, $$

\textcolor{blue}{such that $F_1$ is given by}

$$ F_1 =  \frac{ 2 \cdot TP }{ 2 \cdot TP + FP + FN}. $$

\textcolor{blue}{The relative volume difference is calculated for each of the true positive lesions using
the manual and predicted lesion volumes:}

$$ Relative\,\,volume\,\,difference = \frac{V_{manual} - V_{predicted}}{V_{manual}}. $$

\textcolor{blue}{To illustrate how the performance of our framework varies with lesion size, we calculated
the above measures based on evenly split quantiles of the manual estimates of lesion volumes into 3 groups.  These three size ranges (in terms of the number of voxels) were:  $[3-12)$, $[12-28)$, and $[28-551]$.}

[^4]: \textcolor{blue}{For comparison, the training data set of
the MS Lesion Segmentation challenge
associated with the international MICCAI 2008 conference has a mean lesion load of
204 ($\pm$ 752) mm$^3$ per lesion and the resolution is almost twice what is used
in this study (i.e., 0.5 $\times$ 0.5 $\times$ 0.5).}

### 2016-06-29 results.Rmd
ENH:  Evaluation results.

@@

<!--
i.e.,

$$ Dice = 2 \frac{\sum_r|S_r \cap T_r|}{\sum_r|S_r| + |T_r] } $$
-->

![Voxelwise comparison with manual delineation of white matter hyperintensities.  On the left are the
calculated Dice values over all white matter hyperintensities.  Note the improvement
in the Dice metric from the employment of the Stage 2 component of the processing pipeline.
(Right) Similar results can be seen by comparing the total lesion load volume between manual
and automated detection strategies.  Although some outliers are found after the Stage 2
processing in a couple subjects, the number of outliers caused by false positives is
decreased significantly with the second stage processing.](Figures/llvAndDice.png)

>>>>

\textcolor{blue}{In Figure 8 we provide the segmentation evaluations derived from the leave-one-out evaluation
of the previously described TBI data.  These performance measures include sensitivity,
positive predictive value, $F_1$ score, and relative volume difference.  The three lesion
size ranges over which these measures are computed to illustrate the variation
in performance with lesion size.  Smaller lesions ($< 12$ voxels) are more difficult
to identify which is why the sensitivity is widely varied over the cohort vs. the largest
set of voxels ($> 28$ voxels).   The first three measures are based on the identification of
entire lesions.  The relative volume difference provides a direct assessment of how accurate
the volumetric estimate is when comparing the manually identified lesions versus the
automatically predicted lesions.}

![Evaluation measures for the leave-one-out protocol of the described protocol in the Methods
section:  (a) sensitivity, (b) positive predictive value, (c) $F_1$ score, and (d) relative
volume difference.   The three quantile ranges include the following lesion volume ranges:
$[3-12)$, $[12-28)$, and $[28-551]$ based on the manually-derived lesion volumes.  These
measures are computed for both Stages.  Overall improvement is seen as the
second Stage RF model is applied for all three quantile ranges.](Figures/evaluationResults.png)

@@

### 2016-06-29 stitched.Rmd
ENH:  Evaluation results.

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{1.2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

>>>>

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
  html_document:
    toc: false

header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---
